A user-extensible and adaptable parser architecture

### 2008-08-16 doc/sgai-2008.tex
Small improvements. Changes to reduce space consumption.

%\showtable{filename}{caption}{label}
\newcommand{\showtable}[3]{
    \begin{table}[hbtp]
        \caption{#2}\label{#3}
        \input{#1}
    \end{table}
}

\newcommand{\tableline}[0]{%
    \noalign{\smallskip}%
    \hline%
    \noalign{\smallskip}%
}

@@

\SVN$Id$
\SVNId{}  \today{}

%XXX DISCUSS COMPLEXITY\@?

@@

It was hoped to reuse an existing parser rather than starting from scratch:
several parsers were considered and rejected because their parsing coverage
was too low, their parsing too inexact, and/or they did not extract
sufficient data.  The effort required to adapt and improve an existing
parser was judged to be greater than the effort to write one, as the
techniques used by the existing parsers severely limited their potential:
some ignored the majority of log lines, parsing specific log lines with
great accuracy, but without any provision for readily parsing new or
similar log lines; others sloppily parsed the majority of log lines, but
were incapable of distinguishing between log lines of the same type, e.g.\
rejections, and extracted only a small portion of the data required.  The
only prior published work on the subject of parsing Postfix log files that
we are aware of is~\cite{log-mail-analyser}, which aims to present
correlated log data in a form suitable for a systems administrator to
search using the myriad of standard Unix text processing utilities already
available.  A full state of the art review is outside the scope of this
paper but will be included in the thesis resulting from this work.

The solution developed is conceptually simple: provide a few generic
functions (\textit{actions\/}), each capable of dealing with an entire
category of log lines (e.g.\ rejections), accompanied by a multitude of
precise patterns (\textit{rules\/}), each of which matches all log lines of
a specific type, and only that type (e.g.\ rejection by a specific DNSBL).
XXX IMPROVE:\@ It is an accepted standard to separate the parsing procedure
from the declarative grammar it operates with.  Part of the novelty here is
in the way that the grammars (rules and actions) are themselves procedures.
The relationship to another procedural framework (ATN) is discussed below.
This architecture is ideally suited to parsing inputs where the input is
not fully understood or does not conform to a fixed grammar: the
architecture warns about unparsed inputs and other errors, but continues on
parsing as best it can, allowing the developer of a new parser to decide
which deficiencies are most important and require attention first, rather
than being forced to fix the first error which arises.

@@

There is some similarity between the architecture and William Wood's
Augmented Transition Networks (ATN)~\cite{atns, nlpip}, a tool used in
Computational Linguistics for creating grammars to parse or generate
sentences.  The resemblance between ATN and the architecture is accidental,
but it is interesting how two different approaches have a similar division
of responsibilities, while having different semantics.

@@

\begin{table}[ht]
    \caption{Similarities with ATN}\label{Similarities with ATN}
    % The \smallskip{} below stop the text hitting the lines.
    \begin{tabular}[]{lll}
        \hline
        \noalign{\smallskip}
        ATN           & Architecture    & Similarity                          \\
        \noalign{\smallskip}
        \hline
        \noalign{\smallskip}
        Networks      & Framework       & Determines the sequence of
                                          transitions or actions which        \\
                      &                 & constitutes a valid input.          \\
        Transitions   & Actions         & Save data and impose conditions the
                                          input must meet to be               \\
                      &                 & considered valid.                   \\
        Abbreviations & Rules           & Responsible for classifying input.  \\
        \noalign{\smallskip}
        \hline
        \noalign{\smallskip}
    \end{tabular}
\end{table}

@@

Each action contains the code required to deal with a particular type of
log lines, e.g.\ rejections.  The actions are parser-specific: each parser
author will need to write the actions required from scratch, unless
extending an existing parser.  In the Postfix log parser developed for this
project there are \numberOFactions{} actions and \numberOFrules{} rules,
with a very skewed distribution of rules to actions as shown in
\graphref{Distribution of rules per action}.  The distribution is skewed
for the same reason the architecture is so successful in enabling users to
parse new log lines: most new log lines can be parsed by a combination of a
new rule and an existing action.  It is expected that parsers utilising
this architecture will have a high ratio of rules to actions, hopefully
resulting in simpler rules and clearer distinctions between log lines.
\showgraph{build/graph-action-distribution}{Distribution of rules per
action}{Distribution of rules per action} Unsurprisingly the action with
the highest number of associated rules is \texttt{DELIVERY\_REJECTED}, the
action which handles Postfix rejecting a mail; it is followed by
\texttt{SAVE\_DATA}, the action responsible for handling informative
inputs: these are log lines that supplement the data gathered from other
log lines, but that do not by themselves represent interesting events.  The
third most common action is, perhaps surprisingly, \texttt{UNINTERESTING}:
this action does nothing when executed, allowing uninteresting log lines to
be parsed without causing any effects (it does not imply that the input is
ungrammatical or unparsed).  Generally rules specifying the
\texttt{UNINTERESTING} action parse log lines which are not associated with
a specific mail, e.g.\ warnings about configuration files changing.  The
framework warns about every unparsed log line, to alert the user that they
need to alter or extend their ruleset; the Postfix log parser successfully
parses every input line in the \numberOFlogFILESall{} log files it is
currently being tested with.  The remaining actions have only one or two
associated rules: in some cases this is because there will only ever be one
type of log line for that action, e.g.\ all log lines showing that a remote
client has connected are matched by a single rule and handled by the
\texttt{CONNECT} action; in other cases the action is required to address a
deficiency in the log files, or a complication which arises during parsing.
Taking the \texttt{CONNECT} action as an example, it creates a new data
structure for each client connecting, and saves the data extracted by the
rule (in memory, subsequently to be entered into the database).  If the
data structure already exists it is treated as a symptom of a bug, and the
action issues a warning containing the full contents of the data structure,
plus the log line which has just been parsed.

@@

Storing the rules separately to the parser allows external tools to be
written to detect overlapping rules.  Traditional regexes are equivalent in
computational power to \FA{}; it is possible to convert between regexes and
\FA{}, so regex overlap could be detected by finding a non-empty
intersection of two \FA{}\@.  One formula for \FA{} intersection is given
in~\cite{intersection-of-NFA-using-Z}: $FA1 \cap{} FA2 =
\overline{(\overline{FA1} \cup{} \overline{FA2})}$.  Perl 5.10 regexes are
more powerful than traditional regexes, e.g.\ it is possible to match
correctly nested and balanced brackets to an arbitrary depth:
\newline{}\tab{}\verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!\newline{}
Perl 5.10 regexes can maintain an arbitrary state stack and are thus
equivalent in computational power to \PDA{}, so overlap detection may
require calculating the intersection of two \PDA{}\@.  \CFL{} are
equivalent in computational power to \PDA{}, and intersection of two
\CFL{s} is not closed (i.e.\ intersection can produce a language which
cannot be parsed by a \CFL{}), e.g.\ $a^{*}b^{n}c^{n} \cap a^{n}b^{n}c^{*}
\rightarrow a^{n}b^{n}c^{n}$, so \PDA{} intersection may not be meaningful
in some cases.  Detecting overlap between $n$ regexes requires
$\frac{n(n-1)}{2}$ intersections to be calculated, giving $O(n^2x)$
complexity, where $O(x)$ is the complexity of calculating intersection.
This is certainly not a task to be performed on every parsing run:
detecting overlap amongst the Postfix log parser's \numberOFrules{} rules
would require calculating \numberOFruleINTERSECTIONS{} intersections.

@@

% Do not reformat this!
\begin{table}[htbp]
    \caption{Example rule}
    \empty{}\label{Example rule}
    \begin{tabular}[]{ll}
        \tableline{}%
        Attribute                 & Value                                            \\
        \tableline{}%
        %program                   & \daemon{smtpd}                                   \\
        regex                     & \verb!^connect from (__HOSTNAME__)\[(__IP__)\]$! \\
        action                    & CONNECT                                          \\
        %priority                  & 0                                                \\
        \tableline{}%
    \end{tabular}
\end{table}

@@

    \item [Matching rules to log lines is simple:]  The first matching rule
        wins: there is no backtracking to try alternate rules, no attempt
        is made to pick a \textit{best\/} rule; the first rule which
        matches the line is the rule which parses it and determines the
        action that will be invoked.

@@

\begin{table}[htbp]
    \caption{Computer used to generate statistics}
    \empty{}\label{Computer used to generate statistics}
    \begin{tabular}[]{ll}
        \tableline{}
        Component type  & Component in use                                  \\
        \tableline{}
        CPU             & One dual core 2.40GHz Intel\textregistered{}
                            Core\texttrademark{}2 CPU,                      \\
                        & with 32KB L1 cache and 4MB L2 cache.              \\
        RAM             & 2GB 667 MHz DDR RAM\@.                            \\
        Hard disk       & One Seagate Barracuda 7200 RPM 250GB SATA disk.   \\
        \tableline{}
    \end{tabular}
\end{table}

@@

% XXX MAYBE ADD A TABLE SHOWING MEAN AND STDDEV OF THE RATIOS?

@@

How any architecture scales as the number of rules increases is important,
but it is particularly important in this architecture because it is
expected that there will be a large number of rules.  There are
\numberOFrules{} rules in the full Postfix log parser ruleset, whereas the
minimum number of rules required to parse the \numberOFlogFILES{} log files
is \numberOFrulesMINIMUM{}, \numberOFrulesMINIMUMpercentage{} of the full
ruleset.  A second set of statistics was generated using the minimum
ruleset and the parsing times compared to those generated using the full
ruleset: the percentage parsing time increase when using the full ruleset
instead of the minimal ruleset for optimal, shuffled and reversed orderings
is shown in \graphref{Percentage parsing time increase of maximum ruleset
over minimum ruleset}.
\showgraph{build/graph-full-ruleset-vs-minimum-ruleset}{Percentage parsing
time increase of maximum ruleset over minimum ruleset}{Percentage parsing
time increase of maximum ruleset over minimum ruleset}

It is clear from \graphref{Percentage parsing time increase of maximum
ruleset over minimum ruleset} that the increased number of rules has a
noticeable performance impact with reverse ordering, and a lesser impact
with shuffled ordering.  The optimal ordering shows a mean increase of
\input{build/include-full-ruleset-vs-minimum-ruleset} in parsing time for a
\numberOFrulesMAXIMUMpercentage{} increase in the number of rules.  These
results show that the architecture scales very well as the number of rules
increases, and that sorting the rules is an important optimisation enabling
this.

>>>>

%\showtable{filename}{caption}{label}
\newcommand{\showtable}[3]{
    \begin{table}[btp]
        \caption{#2}\label{#3}
        \input{#1}
    \end{table}
}

\newcommand{\tabletopline}[0]{%
    \hline%
    \noalign{\smallskip}%
}

\newcommand{\tablebottomline}[0]{%
    \noalign{\smallskip}%
    \hline%
}

\newcommand{\tablemiddleline}[0]{%
    \noalign{\smallskip}%
    \hline%
    \noalign{\smallskip}%
}

@@

%\SVN$Id$
%\SVNId{}  \today{}

@@

It was hoped to reuse an existing parser rather than starting from scratch:
several parsers were considered and rejected because their parsing coverage
was too low, their parsing too inexact, and/or they did not extract
sufficient data.  The effort required to adapt and improve an existing
parser was judged to be greater than the effort to write one, as the
techniques used by the existing parsers severely limited their potential:
some ignored the majority of log lines, parsing specific log lines with
great accuracy, but without any provision for readily parsing new or
similar log lines; others sloppily parsed the majority of log lines, but
were incapable of distinguishing between log lines of the same type, e.g.\
rejections, and extracted only a small portion of the data required.  The
only prior published work on the subject of parsing Postfix log files that
we are aware of is~\cite{log-mail-analyser}, which aims to present
correlated log data in a form suitable for a systems administrator to
search using the myriad of standard Unix text processing utilities already
available.  A full state of the art review is outside the scope of this
paper but will be included in the thesis resulting from this work.
The relationship to another procedural framework (ATN) is discussed below.

The solution developed is conceptually simple: provide a few generic
functions (\textit{actions\/}), each capable of dealing with an entire
category of log lines (e.g.\ rejecting a mail), accompanied by a multitude
of precise patterns (\textit{rules\/}), each of which matches all log lines
of a specific type, and only that type (e.g.\ rejection by a specific
DNSBL).  It is an accepted standard to separate the parsing procedure from
the declarative grammar it operates with.  Part of the novelty here is in
the way that the grammar is itself partially procedural (each action is a
separate procedure).  This architecture is ideally suited to parsing inputs
where the input is not fully understood or does not conform to a fixed
grammar: the architecture warns about unparsed inputs and other errors, but
continues on parsing as best it can, allowing the developer of a new parser
to decide which deficiencies are most important and require attention
first, rather than being forced to fix the first error which arises.

@@

There is some similarity (shown in \tableref{Similarities with ATN})
between the architecture and William Wood's Augmented Transition Networks
(ATN)~\cite{atns, nlpip}, a tool used in Computational Linguistics for
creating grammars to parse or generate sentences.  The resemblance between
ATN and the architecture is accidental, but it is interesting how two
different approaches have a similar division of responsibilities, while
having different semantics.

@@

\begin{table}[hbtp]
    \caption{Similarities with ATN}\label{Similarities with ATN}
    % The \smallskip{} below stop the text hitting the lines.
    \begin{tabular}[]{lll}
        \hline
        \noalign{\smallskip}
        ATN           & Architecture    & Similarity                          \\
        \noalign{\smallskip}
        \hline
        \noalign{\smallskip}
        Networks      & Framework       & Determines the sequence of
                                          transitions or actions which        \\
                      &                 & constitutes a valid input.          \\
        Transitions   & Actions         & Save data and impose conditions the
                                          input must meet to be               \\
                      &                 & considered valid.                   \\
        Abbreviations & Rules           & Responsible for classifying input.  \\
        \noalign{\smallskip}
        \hline
        \noalign{\smallskip}
    \end{tabular}
\end{table}

@@

Each action contains the code required to deal with a particular type of
log lines, e.g.\ rejections.  The actions are parser-specific: each parser
author will need to write the actions required from scratch, unless
extending an existing parser.  In the Postfix log parser developed for this
project there are \numberOFactions{} actions and \numberOFrules{} rules,
with a very skewed distribution of rules to actions as shown in
\graphref{Distribution of rules per action}.  The distribution is skewed
for the same reason the architecture is so successful in enabling users to
parse new log lines: most new log lines can be parsed by a combination of a
new rule and an existing action.  It is expected that parsers utilising
this architecture will have a high ratio of rules to actions, hopefully
resulting in simpler rules and clearer distinctions between log lines.
\showgraph{build/graph-action-distribution}{Distribution of rules per
action}{Distribution of rules per action} Unsurprisingly the action with
the highest number of associated rules is \texttt{DELIVERY\_REJECTED}, the
action which handles Postfix rejecting a mail; it is followed by
\texttt{SAVE\_DATA}, the action responsible for handling informative
inputs: these are log lines that supplement the data gathered from other
log lines, but that do not by themselves represent interesting events.  The
third most common action is, perhaps surprisingly, \texttt{UNINTERESTING}:
this action does nothing when executed, allowing uninteresting log lines to
be parsed without causing any effects (it does not imply that the input is
ungrammatical or unparsed).  Generally rules specifying the
\texttt{UNINTERESTING} action are rules to parse log lines which are not
associated with a specific mail, e.g.\ warnings about configuration files
changing.  The framework warns about every unparsed log line, to alert the
user that they need to alter or extend their ruleset; the Postfix log
parser successfully parses every input line in the \numberOFlogFILESall{}
log files it is currently being tested with.  The remaining actions have
only one or two associated rules: in some cases this is because there will
only ever be one type of log line for that action, e.g.\ all log lines
showing that a remote client has connected are matched by a single rule and
handled by the \texttt{CONNECT} action; in other cases the action is
required to address a deficiency in the log files, or a complication which
arises during parsing.  Taking the \texttt{CONNECT} action as an example,
it creates a new data structure for each client connecting, and saves the
data extracted by the rule (in memory, subsequently to be entered into the
database).  If the data structure already exists it is treated as a symptom
of a bug, and the action issues a warning containing the full contents of
the data structure, plus the log line which has just been parsed.

@@

Storing the rules separately to the parser allows external tools to be
written to detect overlapping rules.  Traditional regexes are equivalent in
computational power to \FA{}; it is possible to convert between regexes and
\FA{}, so regex overlap could be detected by finding a non-empty
intersection of two \FA{}\@.  One formula for \FA{} intersection is given
in~\cite{intersection-of-NFA-using-Z}: $FA1 \cap{} FA2 =
\overline{(\overline{FA1} \cup{} \overline{FA2})}$.  Perl 5.10 regexes are
more powerful than traditional regexes, e.g.\ it is possible to match
correctly balanced brackets nested to an arbitrary depth:
\newline{}\tab{}\verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!\newline{}
Perl 5.10 regexes can maintain an arbitrary state stack and are thus
equivalent in computational power to \PDA{}, so overlap detection may
require calculating the intersection of two \PDA{}\@.  \CFL{} are
equivalent in computational power to \PDA{}, and intersection of two
\CFL{s} is not closed (i.e.\ intersection can produce a language which
cannot be parsed by a \CFL{}), e.g.\ $a^{*}b^{n}c^{n} \cap a^{n}b^{n}c^{*}
\rightarrow a^{n}b^{n}c^{n}$, so \PDA{} intersection may not be meaningful
in some cases.  Detecting overlap between $n$ regexes requires
$\frac{n(n-1)}{2}$ intersections to be calculated, giving $O(n^2x)$
complexity, where $O(x)$ is the complexity of calculating intersection.
This is certainly not a task to be performed on every parsing run:
detecting overlap amongst the Postfix log parser's \numberOFrules{} rules
would require calculating \numberOFruleINTERSECTIONS{} intersections.

@@

% Do not reformat this!
\begin{table}[htbp]
    \caption{Example rule}
    \empty{}\label{Example rule}
    \begin{tabular}[]{ll}
        \tabletopline{}%
        Attribute                 & Value                                            \\
        \tablemiddleline{}%
        %program                   & \daemon{smtpd}                                   \\
        regex                     & \verb!^connect from (__HOSTNAME__)\[(__IP__)\]$! \\
        action                    & CONNECT                                          \\
        %priority                  & 0                                                \\
        \tablebottomline{}%
    \end{tabular}
\end{table}

@@

    \item [Matching rules to log lines is simple:]  The first matching rule
        determines the action that will be invoked: there is no
        backtracking to try alternate rules, no attempt is made to pick a
        \textit{best\/} rule.

@@

% XXX MAYBE CHANGE POSITIONING
\begin{table}[htbp]
    \caption{Computer used to generate statistics}
    \empty{}\label{Computer used to generate statistics}
    \begin{tabular}[]{ll}
        \tabletopline{}
        Component  & Component used                                 \\
        \tablemiddleline{}
        CPU        & 1 dual core 2.40GHz Intel\textregistered{}
                     Core\texttrademark{}2 CPU,                      
                     with 32KB L1 and 4MB L2 cache.                 \\
        RAM        & 2GB 667 MHz DDR RAM\@.                         \\
        Hard disk  & 1 Seagate Barracuda 7200 RPM 250GB SATA disk.  \\
        \tablebottomline{}
    \end{tabular}
\end{table}

@@

@@

\newpage{} % XXX
How any architecture scales as the number of rules increases is important,
but it is particularly important in this architecture because it is
expected that there will be a large number of rules.  There are
\numberOFrules{} rules in the full Postfix log parser ruleset, whereas the
minimum number of rules required to parse the \numberOFlogFILES{} log files
is \numberOFrulesMINIMUM{}, \numberOFrulesMINIMUMpercentage{} of the full
ruleset.  A second set of statistics was generated using the minimum
ruleset and the parsing times compared to those generated using the full
ruleset: the percentage parsing time increase when using the full ruleset
instead of the minimal ruleset for optimal, shuffled and reversed orderings
is shown in \graphref{Percentage parsing time increase of maximum ruleset
over minimum ruleset}.  It is clear from \graphref{Percentage parsing time
increase of maximum ruleset over minimum ruleset} that the increased number
of rules has a noticeable performance impact with reverse ordering, and a
lesser impact with shuffled ordering.  The optimal ordering shows a mean
increase of \input{build/include-full-ruleset-vs-minimum-ruleset} in
parsing time for a \numberOFrulesMAXIMUMpercentage{} increase in the number
of rules.  These results show that the architecture scales very well as the
number of rules increases, and that sorting the rules is an important
optimisation enabling this.
\showgraph{build/graph-full-ruleset-vs-minimum-ruleset}{Percentage parsing
time increase of maximum ruleset over minimum ruleset}{Percentage parsing
time increase of maximum ruleset over minimum ruleset}
